---
title: "36-315 Final Project"
author: "Alva He, Jialing Deng, Ruonan Sun, Vincent Huai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: show
  pdf_document:
    toc: yes
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE)

```

# Packages and Data

```{r, results = FALSE}

# Libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggridges))
suppressPackageStartupMessages(library(dendextend))
suppressPackageStartupMessages(library(factoextra))
suppressPackageStartupMessages(library(ggcorrplot))
suppressPackageStartupMessages(library(GGally))

# Data
sleep <- read.csv("cmu-sleep.csv")
sleep$cat_race <- as.factor(sleep$demo_race)
sleep$cat_gender <- as.factor(sleep$demo_gender)
sleep$cat_firstgen <- as.factor(sleep$demo_firstgen)

# Data (nh Cohort Excluded)
# nh cohort lacks term_units and Zterm_units_ZofZ
# sleep_nh <- sleep %>%
#   filter(!(cohort == "nh"))

# Data (NA Filtered)
# sleep_clean <- na.omit(sleep)

# Sleep and GPA Variables
sleep_quant <- sleep %>%
  dplyr::select(bedtime_mssd:term_units) %>%
  na.omit()

# Sleep, GPA and Term Variables
sleep_quantextra <- sleep %>%
  dplyr::select(bedtime_mssd:Zterm_units_ZofZ) %>%
  na.omit()

```

# Introduction

## Motivation

It is important for students to succeed during their first year of college for numerous reasons. However, the transition to college life presents many challenges, which can often result in the comprising of a student's sleep. Sleep is crucial to cognitive function, so the reduction in sleep could threaten a student's ability to suceed in their first year of college. We hypothesize that better sleep could lead to a higher GPA, and suggest that university policy and student behavior can be adjusted accordingly to enhance academic outcomes.

## Overall Theme

The overarching aim is to discern the relationship between sleep habits and academic success during the pivotal first year of college. This exploration is structured into three distinct sections: 
- Firstly, the connection between sleep duration and academic achievement is examined. 
- Secondly, we consider whether sleep variability, as an indicator of sleep quality, might also correlate with GPA. 
- Lastly, we identify if these relationships hold true when accounting for demographic factors like race, gender, and first-generation college status.

# Overview

Our data was collected from a study on the CMU Data Repository that surveyed first-year students from Carnegie Mellon University (CMU), The University of Washington (UW) and Notre Dame University (ND). In total, 634 students participated in the survey Students received a Fitbit to track their sleep and physical activity. Additionally, their GPAs were collected from their university's registrar. 

Variables such as subject ID, study number, cohort, and demographic details are featured, alongside sleep-related metrics like bedtime variability and total sleep time. The investigation delves into the potential influence of sleep duration on academic performance, specifically changes in the end-of-semester grade point average (GPA) among first-year college students.

We have five categorical variables and ten quantitative variables as displayed below. Some variables described various features of the student such as race and gender; some variables described the sleeping habits of the student; and some variables described the student's academic performance.

| Descriptive Variable | Description |
|:--- |:--- |
| *Subject ID* | Unique ID of the Subject. |
| *Study* | Study Number (corresponding to last table). |
| *Cohort* | Codename of the cohort that the subject belongs to. |
| *Race* | Binary label for underrepresented and non-underrepresented students (underrepresented = 0, non-underpresented = 1). |
| *Gender* | Gender of the subject (male = 0, female = 1), as reported by their institution. |
| *First Generation* | First-generation status (non-first gen = 0, first-gen = 1). |

| Sleep-Related Metric | Description |
|:--- |:--- |
| *Bedtime MSSD* | Mean successive squared difference of bedtime. This measures bedtime variability, and is calculated as the average of the squared difference of bedtime on consecutive nights. |
| *Total Sleep Time* | Average time in bed (the difference between wake time and bedtime) minus the length of total awake/restlessness in the main sleep episode, in minutes. |
| *Midpoint Sleep* | Average midpoint of bedtime and wake time, in minutes after 11 pm. |
| *Fraction Nights with Data* | Fraction of nights with captured data for the subject. |
| *Daytime Sleep* | Average sleep time outside of the range of the main sleep episode, in minutes. |

| Academic Performance Metric | Description |
|:--- |:--- |
| *Cumulative GPA* | Cumulative GPA (out of 4.0), for previous semesters. |
| *Term GPA* | End-of-term GPA (out of 4.0) for the current semester. |
| *Term Units* | Number of course units carried in the term. |
| *Term Units (Adjusted)* | Term Units adjusted for mean of 0 and standard deviation of 1. |


| Study | University | Semester |
|:--- |:--- |:--- |
| 1 | Carnegie Mellon University | Spring 2018 |
| 2 | University of Washington | Spring 2018 |
| 3 | University of Washington | Spring 2019 |
| 4 | Notre Dame University | Spring 2016 |
| 5 | Carnegie Mellon University | Spring 2017 |

# Research Questions

We investigated the following three main research questions in our analysis.

\vspace{10pt}
**1. What is the impact of sleep patterns on GPA?**

We hypothesize that students with irregular or insufficient sleep may experience fluctuations in their term GPA. 


\vspace{10pt}
**2. What is the impact of being a first-generation or underrepresented student on academic outcomes? Does gender play a role in academic success?**


\vspace{10pt}
**3. Do students from different schools experience different levels of sleep or academic performance?**



# Sleep Patterns and GPA

Through a series of exploratory data analyses (EDA), including scatterplots, pairs plot and histograms, we explore trends and quantify the relationship between total sleep time and GPA. Furthermore, advanced visualizations like PCA biplots and correlograms are be employed to understand the multidimensional interactions among various sleep-related metrics. 

This approach will provide insight into any linear or more complex relationship influencing academic success.


We explore the relationships between sleep and academic success with the following graphs.
- Histograms
- Correlogram
- Principal Component Analysis (PCA)
- Scatterplots
- Heat maps
- Pairs plots

## Histograms

In the following code chunk, we plot histograms and density plots to explore the distribution of important variables such as `TotalSleepTime`, and `term_gpa`.

```{r}

sleep %>%
  ggplot2::ggplot(aes(x = TotalSleepTime)) +
  ggplot2::labs(title = "Total Sleep Time Distribution",
                y = "Density",
                x = "Total Sleep Time (minutes)") +
  ggplot2::geom_histogram(aes(y = ggplot2::after_stat(density)),
                          color = "deepskyblue4", 
                          fill = "deepskyblue",
                          binwidth = 11.59) +
  ggplot2::geom_density(fill = "deeppink",
                        alpha = 0.2)

```

Sleep Time appears to be approximately normally distributed, centered around 400 minutes (which is roughly 6.5 hours). There are some outliers on both the lower and higher ends of the sleep time, but the bulk of the data falls within the normal range. The corresponding density plot confirms the bell-shaped curve. 

```{r}

sleep %>%
  ggplot2::ggplot(aes(x = term_gpa)) +
  ggplot2::labs(title = "Term GPA Distribution",
                y = "Density",
                x = "Term GPA") +
  ggplot2::geom_histogram(aes(y = ggplot2::after_stat(density)),
                          color = "darkorchid4", 
                          fill = "darkorchid1",
                          binwidth = 0.1066) +
  ggplot2::geom_density(fill = "green",
                        alpha = 0.2)

```

The histogram for Term GPA shows that the data skews left, indicating that more students have a GPA closer to 4.0 than to the lower end of the scale. There's a clear peak around the GPA of 3.5. The density plot overlays the histogram, providing a smoothed curve representation of the distribution, emphasizing the skew towards higher GPAs.

## Correlogram

```{r}

sleep_quant.cor <- cor(sleep_quant)

sleep_quant.cor %>%
  ggcorrplot::ggcorrplot(method = "circle", hc.order = FALSE, type = "lower") +
  ggplot2::ggtitle("Correlogram for Sleep and GPA variables")

```

## PCA Biplot

The intuition of making PCA biplot is to identify any linear relationships between sleeping time and cumulative GPA. We aim to investigate how different aspects of sleep relate to academic performance.

```{r}

# Create PCA
sleep_pca <- prcomp(sleep_quantextra, center = TRUE, scale. = TRUE)
summary(sleep_pca)

```

```{r}

# Create PCA matrix
sleep_pca_matrix <- sleep_pca$x
head(sleep_pca_matrix)

```

```{r}

sleep_pca %>%
  fviz_pca_biplot(label = "var",
                  alpha.ind = 0.25,
                  alpha.var = 0.75,
                  col.var = "darkblue",
                  repel = TRUE)

```

The correlation between `TotalSleepTime` and `cum_gpa` is weakly positive. This implies that more total sleep might be associated with higher GPA. 

The correlation between `midpoint_sleep` and `cum_gpa` is negative. This suggests that later midpoints of sleep are slightly associated with lower GPA. The PCA provides a potential further research question: Given the weak positive correlation, it would be important to determine if longer sleep directly contributes to higher GPA or if they are both influenced by a third factor.

## Scatterplots

My motivation for this plot is to explore the relationship between students' total sleep time and their academic performance, classified by their term GPA, with a further consideration of the average midpoint of their sleep. This is informative because this visualization underscores the importance of not only the length of sleep but also the timing of sleep in relation to academic performance. Whether staying up late might have different effects for the same length of sleep.

```{r}

sleep <- sleep %>%
  mutate(term_grade = ifelse(term_gpa < 2.50, "At Risk",
                             ifelse(term_gpa < 3.00, "Good",
                                    ifelse(term_gpa < 3.50, "Very Good", 
                                           "Excellent"))))

sleep %>%
  ggplot(aes(x = midpoint_sleep, y = TotalSleepTime, color = term_grade)) + 
  labs(title = "Total Sleep Time vs. Average Midpoint Sleep by GPA",
       y = "Total Sleep Time (minutes)",
       x = "Average Midpoint Sleep",
       color = "Term Grade") +
  geom_point(alpha = 0.3) + 
  geom_smooth(method = "lm", se = FALSE)

```

For students with good, very good, and excellent term grades, there appear negative relationships for the average midpoint of sleep and the total length of sleep, which indicates that these students tend to reduce their sleep length to make up for staying up late. It is worth noting that students with excellent and very good term grades tend to have greater length of sleep than students with good term grades. However, for students with At Risk term grade, there appears a positive relationship between the average midpoint of sleep and the total length of sleep, which indicates that the later the students sleep, the longer time they sleep in the main sleep episode. Also, for the earlier average midpoint of sleep (earlier sleep time), students with higher term gpa tend to sleep longer, but for the later average midpoint of sleep (later sleep time), students with lower term gpa tend to sleep longer. 

### Regression Analysis

##Model1: pure sleep-related data

```{r}

sleep_lm <- lm(term_gpa ~ bedtime_mssd + TotalSleepTime + midpoint_sleep
               + frac_nights_with_data + daytime_sleep, data = sleep)
summary(sleep_lm)

```

**TotalSleepTime:** The coefficient $0.0017695$ is positive and statistically significant with a p-value of $0.000692$. This suggests that an increase in Total Sleep Time is associated with an increase in GPA, controlling for other factors.

**midpoint_sleep:** The coefficient $-0.0009123$ is negative and significant with a p-value of $0.010950$. This indicates that a later midpoint of sleep (which could indicate going to bed later) is associated with a lower GPA.

**daytime_sleep:** The coefficient $-0.0027385$ is negative and significant with a p-value of $0.002823$. THis suggests that more sleep during the day is associated with a lower GPA.

**bedtime_mssd and frac_nights_with_data:** These coefficients are not statistically significant (their p-values are $0.107901$ and $0.391353$ respectively). This indicates that there is no clear evidence that variability in bedtime and the fraction of nights with data are associated with GPA within this model.

**Adjusted R-Squared:** The R-squared value of $0.08109$ indicates that the model explains about $8.1%$ of the variability in the dependent variable. This is relatively low, which means that there are other factors other than sleep time that are not included in the model that affect GPA.

```{r}

qqnorm(resid(sleep_lm))
qqline(resid(sleep_lm))

```

**Residuals vs Fitted:**
The plot shows a random pattern of residuals, which is good for the assumption of homoscedasticity. However, there may be a slight curve, indicating potential non-linearity.
Normal Q-Q Plot:
There's some deviation from the normal line, particularly in the tails, which suggests that the residuals may not be normally distributed and there might be outliers or influential points affecting the model.

**Residuals vs Leverage:**
This plot shows that there are a few points with higher leverage, but they do not seem to have a significant influence on the model, as indicated by the Cook's distance not exceeding the threshold.

#Model2: total sleep time as the representative for regression data, control for previous term GPA,cohort, and term units 

```{r}
sleep$study <- as.factor(sleep$study)

# Now rename the levels
levels(sleep$study) <- c( "1", "2", "3", "5")

# Now we can fit the model with the renamed factor levels
model2 <- lm(term_gpa ~ TotalSleepTime + cum_gpa + study + term_units, data = sleep)

# Summary of the model
summary(model2)

# Diagnostic plots
par(mfrow=c(2,2))
plot(model2)
```
**TotalSleepTime:** The coefficient $0.0018722$ is positive and statistically significant, which suggests that an increase in Total Sleep Time is associated with an increase in GPA, controlling for other factors.
**cum_gpa:** we control for cum_gpa because previous semesters’ cumulative gpa is highly correlated with the new semester’s GPA, which reflects the underlying studying ability of the students. It can be clearly seen from the model that they are highly positively correlated. 

**study**: the cohort effect is not statistically significant, indicating that there is no difference in sleep patterns vs GPA for different studies.(We can make generalization ) 

**term_units:** The coefficient is $0.0018998$ and is not statistically significant (p-value = $0.711$), suggesting that the number of term units is not a significant predictor of term GPA in this model.

**Adjusted R-Squared:** The value of $0.4106$ indicates that the model explains about 41.6%of the variability in the dependent variable, and the contribution can be largely attributed to previous terms’ cumulative GPA
**F-Statistic:** The F-statistic is significant, indicating that the model is jointly significant. 


## Pairs Plot

```{r, warning = FALSE}

# Create a pairs plot using GGally
sleep %>%
  ggpairs(columns = c("bedtime_mssd", "TotalSleepTime", "midpoint_sleep", "term_gpa"),
          mapping = aes(color = cat_gender, shape = cat_race, alpha = cat_firstgen),
          title = "Pairs Plot of Sleep and GPA Variables")

```

Intuition: By stratifying the data with colors and shapes based on demographic factors like gender and race, and using transparency to differentiate between first-gen and non-first-gen students, the plot can reveal whether these subgroups experience different relationships between sleep and academic performance.

Inconsistent relationship when only look at quantitative variable vs looking at categorical and quantitative together→ The relationship between sleep time and GPA could be non-linear. Perhaps both too little and too much sleep could be detrimental to GPA, which might result in a U-shaped relationship that would not be well-captured by a simple linear correlation.

# Demographics and GPA


Initial observations indicate that disparities in academic outcomes may correlate with demographic variables such as race, gender, and first-generation college status. 

This part of the study focuses on examining how these factors influence term GPA among students. By utilizing histograms to visualize GPA distributions and conducting one-way ANOVA tests, this research will evaluate differences across demographic groups. 

This statistical approach will help determine if demographic factors serve as significant predictors of academic achievement, potentially informing targeted support strategies for underrepresented groups.


-Histogram on the distribution of term_gpa (EDA)
-one-way ANOVA analysis to see whether term_gpa differs across different demographic groups (gender, race, and first-generation status).

```{r}
library(tidyverse)
sleep<-read.csv("cmu-sleep.csv")
library(ggplot2)

sleep$demo_firstgen = as.factor(sleep$demo_firstgen)
sleep$demo_race = as.factor(sleep$demo_race)
sleep$demo_gender = as.factor(sleep$demo_gender)

sleep_clean <- sleep |>
  filter(demo_firstgen != "2") |>
  drop_na(term_gpa, demo_gender, demo_firstgen, demo_race) |>
  mutate(
    demo_firstgen = as.factor(demo_firstgen),
    demo_race = as.factor(demo_race),
    demo_gender = as.factor(demo_gender)
  )

#term gpa vs. first-gen status
sleep_clean |>
  ggplot(aes(x = term_gpa, color =demo_firstgen)) +
  geom_density() +
  scale_color_discrete(labels = c("non-first gen", "first-gen")) +
  labs(x = "Term GPA", y = "Density", color = "First-generation status",
       title = "Term GPA with first-generation status") 

t.test(term_gpa ~ demo_firstgen, data = sleep_clean)

#term gpa vs. gender status
sleep_clean |>
  ggplot(aes(x = term_gpa, color =demo_gender)) +
  geom_density() +
  scale_color_discrete(labels = c("male", "female")) +
  labs(x = "Term GPA", y = "Density", color = "Gender",
       title = "Term GPA with gender status")  

t.test(term_gpa ~ demo_gender, data = sleep_clean)

#term gpa vs. race status
sleep_clean |>
  ggplot(aes(x = term_gpa, color =demo_race)) +
  geom_density() +
  scale_color_discrete(labels = c("underrepresented", "non-underrepresented")) +
  labs(x = "Term GPA", y = "Density", color = "Race",
       title = "Term GPA with race status") 

t.test(term_gpa ~ demo_race, data = sleep_clean)
```

Interpretation for first-generation status vs. term gpa:
The density plot illustrates the distribution of term GPA scores for two distinct groups: first-generation college students ("first-gen") and non-first-generation students ("non-first gen"). In general trend, the first-gen student distribution has its peak slightly to the left of the non-first-gen distribution. This suggests a higher concentration of first-gen students within the lower GPA range. Conversely, the non-first-gen students display a somewhat higher density towards the upper end of the GPA scale, indicating a tendency for these students to achieve higher GPAs. Despite these differences, the overall shapes of both distributions are similar, peaking around a GPA of 3, indicating that the most common GPA scores for both first-gen and non-first-gen students. From the t-test, we observed that the mean GPA for non-first-gen students is approximately 3.4803, and the mean for first-gen students is approximately 3.2966. Since the p-value is smaller than 0.05,  we conclude that non-first-gen students have a higher average GPA.

Interpretation for gender status vs. term gpa:
The density plot illustrates the distribution of term GPA scores for male and female students, with both distributions peaking around a GPA of 3.The corresponding t-test shows that the mean GPA for male students is about 3.45219 and female students about 3.448530. Since the p-value for the t-test is greater than 0.05, there's no statistically significant difference in mean GPA between male and female students. 

Interpretation for race status vs. term gpa:
The density plot illustrates the distribution of term GPA scores for underrepresented and non-underrepresented race groups. The distribution of GPAs for the underrepresented group might be slightly shifted towards the lower end of the GPA range compared to the non-underrepresented group, which indicates students with non-underrepresented race tends to perform better for top term GPAs (above 3.5). The t-test results show a t-value of approximately -3.8419 with a p-value of roughly 0.000178, which is well below the typical significance level of 0.05. This indicates a statistically significant difference in the mean GPAs between the two groups, with the non-underrepresented group having a higher average term GPA.


# Cohort and Performance

Preliminary data suggests there might be differences in student lifestyles, such as sleep habits, across various universities. 

This segment aims to explore these differences and identify potential trends or anomalies in student behavior by comparing groups from different institutions (i.e. CMU with others). Techniques such as dendrograms for hierarchical clustering, violin plots for detailed distribution comparisons, and mosaic plots coupled with Chi-square tests will be applied to investigate these variances. 

This analysis will not only highlight how study environments might influence lifestyle choices but also explore the broader impacts of institutional characteristics on student demographics and behaviors.

-Dendrogram by study groups
-Violin plot on TotalSleepTime by study groups
-Mosaic plot, Chi-square test for independence for comparing the number of first-generation by study cohort

## Mosaic Plot

```{r, warning = FALSE}

mosaicplot(table(sleep$demo_firstgen, sleep$study), main="First Generations by studies", shade = TRUE)
chi_square_results <- chisq.test(table(sleep$demo_firstgen, sleep$study))

```

```{r}

chisq.test(table(sleep$demo_firstgen, sleep$study))

```

The mosaic plot shows the standardized residuals from a chi-squared test of independence between two categorical variables: first-generation status (non-first gen = 0, first-gen = 1) and cohort (i.e. which study the subject participate in) . In mosaic plots, the width of the rectangles is proportional to a marginal distribution, and the height is proportional to a conditional distribution. Thus, from the above plots, we can easily assess the marginal distribution of first-gen.The area of each box corresponds to the number of observations in each combination of categories. In our plot, the combination first generation/study1 (CMU 2018) has an observed count that is significantly lower than what would be expected under independence. 
it appears that there are also significantly more “first generations” for study 2 than we would expect under the null hypothesis of independence, thereby suggesting that we should reject this null hypothesis. For study 5, there seems to be fewer first generation participants than we expected. 

We can also verify our observation through conducting chi-square test for independence, and the result shows that we get a p-value close to 0. From both tests, we reject the null hypothesis that whether the student is first generation and study are independent, thereby suggesting that these variables are dependent

## Scatterplots

```{r}

sleep %>%
  ggplot(aes(x = TotalSleepTime, y = cum_gpa)) +
  labs(title = "Cumulative GPA vs. Total Sleep Time",
       y = "Cumulative GPA",
       x = "Total Sleep Time (minutes)",
       color = "Study") +
  geom_point(aes(color = as.factor(study)), alpha = 0.4) +
  geom_smooth(method = "lm", aes(group = as.factor(study), color = as.factor(study)), alpha = 0.3) +
  theme_minimal()

```

Interpretation: There is a positive trend indicated by the trend lines for each study, and the slope is roughly the same, suggesting that as TotalSleepTime increases, cum_gpa tends to increase as well.
The shaded areas around each trend line represent the confidence interval for the trend line's estimate. We see very wide shaded areas indicating relatively high uncertainty about the trend line's slope.

Outliers are present in the data, with some students getting much less or much more sleep than the average. It's interesting to note that at both extremes of sleep duration, there are students with high GPAs, indicating that some students might sacrifice sleep time  for studying; and some who get sample sleep get more focused at class and higher study  efficiency.

## Violin Plot

```{r}

sleep %>%
  ggplot(aes(x = factor(study), y = TotalSleepTime)) +
  geom_violin() +
  geom_boxplot(width = 0.5) +
  geom_point(alpha = 0.6, aes(color = factor(demo_gender))) +
  scale_color_discrete(name = "Gender of the subject", 
                       breaks=c("0", "1"), 
                       labels =c("male", "female")) +
  labs(title = "Violin Plot of Total Sleeping Time by Study Groups", x = "Study Groups", y = "Total Sleeping Time, in minutes") +
  scale_x_discrete(breaks=c("1","2", "3", "4", "5"),
        labels=c("CMU", "UW", "UW", "ND", "CMU"))

```

The questions I am interested in exploring are whether CMU students sleep significantly less than students from other universities and whether female students sleep more than male students. This violin plot shows the full shape of the distribution of total sleeping time regarding students from each university, with each specific data point being categorized by the gender of the investigated students. According to the box plots enclosed in the violin shapes, the average median sleeping time among CMU students is not significantly less than that of students from other universities. According to the density curves given by the violin plot, the peaks of the curves for CMU students' sleeping time are also not significantly lower than those of other universities. We can see that among all universities, most male students tend to have an intermediate amount of sleep while female students are more likely to sleep very little or a lot, except for students from the University of Washington.

## Dendrogram

```{r}

# standardize

# dist matrix
sleep_dist <- sleep_quantextra %>%
  scale(center = FALSE,
        scale = apply(sleep_quantextra, 2, sd, na.rm = TRUE))

sleep_dist <- sleep_dist %>%
  dist(sleep_quant, method = "euclidean")

# plot(as.dendrogram(hclust(sleep_dist, method = "single")))
# plot(as.dendrogram(hclust(sleep_dist, method = "complete")))

```

```{r}

sleep_complete_dend <- as.dendrogram(hclust(sleep_dist, method = "complete"))

```

```{r}

sleep_complete_dend <- set(sleep_complete_dend, "branches_k_color", k=5)

sleep_colors <- ifelse(sleep$study == 5, "red",
                       ifelse(sleep$study == 4, "orange",
                              ifelse(sleep$study == 3, "gold",
                                     ifelse(sleep$study == 2, "green", "blue"))))

plot(set(sleep_complete_dend, "labels_colors",
         order_value = TRUE, sleep_colors))

```

dendrogram (five branches colored, with which study they came from as the label colors)


##Conclusion 

**First-year college students are getting insufficient sleep, and the present work indicates that it may carry significant costs for their academic achievement. Over 600 students in five studies drawn across three different American universities slept on average 6 h and 37 min per night early in their academic terms, a rate highlighting significant accumulating sleep debt. Moreover, every hour of nightly sleep lost was associated with a 0.18% decrease in end-of-term GPA. The present findings call for more research on sleep–achievement outcomes in young adults, and encourage new public health efforts to help young adults get more sleep.**



## Unfinished Work

```{r}

# sleep %>%
#   ggplot() +
#   geom_point(data = subset(sleep, daytime_sleep <= 77), 
#              aes(x = midpoint_sleep, y = TotalSleepTime, color = daytime_sleep), 
#              alpha = 0.85) +
#   scale_color_gradient2("Average \nDaytime Sleep \n(minutes)", 
#                         low = "red", mid = "orange", high = "blue",
#                         midpoint = median(sleep$daytime_sleep)) +
#   geom_point(data = subset(sleep, daytime_sleep > 77),
#              aes(x = midpoint_sleep, y = TotalSleepTime), 
#              alpha = 0.85) +
#   labs(title = "Total Sleep Time vs. Midpoint of Sleep",
#        x = "Midpoint of Sleep (minutes after 11pm)",
#        y = "Total Sleep Time (minutes)")
#   

```

```{r}

# sleep %>%
#   ggplot(aes(x = bedtime_mssd, y = midpoint_sleep, color = Zterm_units_ZofZ)) +
#   scale_color_gradient2(low = "red", mid = "orange", high = "blue") +
#   geom_point(alpha = 0.95)
# 
# sleep %>%
#   ggplot(aes(x = frac_nights_with_data, y = cum_gpa, color = daytime_sleep)) +
#   scale_color_gradient2(low = "red", mid = "orange", high = "blue") +
#   geom_point(alpha = 0.5)

```

```{r}

# sleep %>%
#   ggplot(aes(x = cat_gender, fill = cat_race)) +
#   facet_wrap(~ cat_firstgen) +
#   geom_bar(position = "dodge")

```

```{r}



# sleep %>%
#   ggplot(aes(y = cat_gender, x = cum_gpa)) +
#   facet_grid( ~ cat_race) +
#   ggridges::geom_density_ridges(rel_min_height = 0.01,
#                                 alpha = 0.75,
#                                 aes(fill = cat_firstgen))

```

```{r}

# sleep %>%
#   ggplot(aes(x = daytime_sleep, y = term_gpa, color = Zterm_units_ZofZ)) +
#   geom_point(alpha = 0.75) +
#   scale_color_gradient2("Term Units \n(Z-Score)",
#                         limit = c(-2, 2),
#                         low = "red", mid = "green", high = "blue", 
#                         na.value = rgb(1, 0.96, 0.83, alpha = 0.001)) +
#   labs(title = "Term GPA vs. Daytime Sleep",
#        x = "Daytime Sleep (minutes)",
#        y = "Term GPA") +
#   geom_smooth(method = "lm", se = FALSE)

```

```{r}

# sleep %>%
#   ggplot(aes(x = TotalSleepTime, y = term_gpa, color = midpoint_sleep)) +
#   scale_color_gradient(low = "orange", high = "blue") +
#   geom_point(alpha = 0.5)

```

```{r}

# sleep %>%
#   ggplot(aes(x = Zterm_units_ZofZ, y = term_gpa, color = TotalSleepTime)) +
#   geom_point(alpha = 0.85) +
#   labs(title = "Total Sleep Time vs. Midpoint of Sleep",
#        x = "Term Units (Z-Score)",
#        y = "Term GPA") +
#   geom_smooth(method = "lm", se = FALSE)
# 
# sleep <- sleep %>%
#   subset(daytime_sleep < 250)
# 
# summary(lm(cum_gpa ~ daytime_sleep, data = sleep))

```

```{r}

# filter(sleep, cohort == "nh")$Zterm_units_ZofZ %>%
#   is.na() %>%
#   sum()

```

```{r}

# sleep_pca <- prcomp(sleep_quant, center = TRUE, scale. = TRUE)
# summary(sleep_pca)

```

```{r}

# fviz_eig(sleep_pca, choice = "variance", ncp = 9, addlabels = TRUE) +
#   geom_hline(yintercept = 100 * (1 / ncol(sleep_quant)))

```

```{r}

# sleep_nh <- sleep_nh %>%
#   mutate(pc1 = sleep_pca$x[,1],
#          pc2 = sleep_pca$x[,2],
#          pc3 = sleep_pca$x[,3])
# 
# sleep_nh %>%
#   ggplot(aes(x = pc1, y = pc2)) +
#   labs(title = "PCA Plot for Sleep",
#        x = "PC 1",
#        y = "PC 2") +
#   geom_point(aes(color = as.factor(study)), alpha = 0.75)

```

```{r}

# fviz_pca_biplot(sleep_pca, label = "var",
#                 alpha.ind = 0.25,
#                 alpha.var = 0.75,
#                 repel = TRUE,
#                 # Set the color of the points to decades variable:
#                 habillage = sleep_nh$cohort, pointshape = 19)

```

```{r}

# # standardize
# 
# # dist matrix
# sleep_dist <- sleep_quant %>%
#   scale(center = FALSE,
#         scale = apply(sleep_quant, 2, sd, na.rm = TRUE))
# 
# sleep_dist <- sleep_dist %>%
#   dist(sleep_quant, method = "euclidean")
# 
# plot(as.dendrogram(hclust(sleep_dist, method = "single")))
# plot(as.dendrogram(hclust(sleep_dist, method = "complete")))

```

```{r}

# sleep_complete_dend <- as.dendrogram(hclust(sleep_dist, method = "complete"))

```

```{r}

# sleep_complete_dend <- set(sleep_complete_dend, "branches_k_color", k=5)
# 
# plot(sleep_complete_dend)

```

```{r}

# sleep_colors <- ifelse(sleep$study == 5, "red",
#                        ifelse(sleep$study == 4, "orange",
#                               ifelse(sleep$study == 3, "gold",
#                                      ifelse(sleep$study == 2, "green", "blue"))))
# 
# plot(set(sleep_complete_dend, "labels_colors", 
#          order_value = TRUE, sleep_colors))

```